Open archaeology, open source?

### 2023-10-17 analysis/_04-open_archaeology.qmd
Expand a bit on licenses
Written in haste...

```{r data-licenses}
oarch_licenses <- read_csv(here("analysis/data/raw_data/open-archaeo_licenses.csv"))
oarch_licenses |>
  select(gh_repo = url, other_license = license) |>
  right_join(oarch, by = "gh_repo") |>
  hoist(gh_repo_info, license = list("license", 1, "spdx_id")) |>
  mutate(license = if_else(license == "NOASSERTION", other_license, license)) |>
  select(-other_license) ->
  oarch
```

@@

We found that the projects that host their code on GitHub use a wide variety of standard licenses.
However, nearly half of all repositories <!-- TODO: encode the percentage --> do not explicitly denote an open source license, or any license at all [@tbl-repo-licenses].
Maintainers may have assumed that making the code available, without explicitly stating permitting use, is enough to allow unrestricted access to the repository's contents.
This is evocative of a similar ignorance among academics regarding the unsustainable use of GitHub as a long-term code and data hosting platform (@milliken2021; @escamilla2022; @escamilla2023).
<!-- TODO: Expand with additional info after seeing what is in the "other" category -->

>>>>

```{r data-licenses}
oarch_licenses <- read_csv(here("analysis/data/raw_data/open-archaeo_licenses.csv"))
oarch_licenses |>
  select(gh_repo = url, other_license = license) |>
  right_join(oarch, by = "gh_repo") |>
  hoist(gh_repo_info, license = list("license", 1, "spdx_id")) |>
  mutate(license = if_else(license == "NOASSERTION", other_license, license)) |>
  select(-other_license) ->
  oarch

p_license_gpl <- n_filter(oarch, str_detect(license, "GPL")) / n_filter(oarch, !is.na(license))
p_license_mit <- n_filter(oarch, str_detect(license, "MIT")) / n_filter(oarch, !is.na(license))
p_license_cc <- n_filter(oarch, str_detect(license, "CC")) / n_filter(oarch, !is.na(license))
```

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license [@tbl-repo-licenses].
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restricts they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of FOSS software within non-FOSS software, [@cite?]).
Archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@Balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@cite?].
Many repositories do not specify a license; their maintainers may have assumed that making the code available, without explicitly stating permitting use, is enough to allow unrestricted access to the repository's contents.
This is evocative of a similar ignorance among academics regarding the unsustainable use of GitHub as a long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023].
<!-- TODO: Seems like a pair of slightly uncharitable assumptions? -JR -->
<!-- TODO: (ZB) Expand with additional info after seeing what is in the "other" category -->
